---
title: "baseball"
output: html_notebook
description: create an SQL database using csv files on baseball games, also normalize the database
---
1. read the files

```{r}
library(readr)
game_log <- read_csv("game_log.csv")
park_codes <- read_csv("park_codes.csv")
person_codes <- read_csv("person_codes.csv")
team_codes <- read_csv("team_codes.csv")

```
2. short descriptions
game_log has 161 columns, with detailed inforamtion of each game, including park_id, ids for each player, and team_id
park_codes has park_id and name
person_codes has id, name, when the person was a manager, coach, ump
team_codes has team_id, league, start, end, city, nickname, franch_id, seq

defensive position numbers:
putouts, a fielder that makes someone out
assists, catches the ball before throwing it to someone that gets a putout
errors, missed the ball
passed_balls, catchers who did not catch a legal ball
double plays, had two outs
triple plays, had three outs

league:
UA, Union Association (old and extinct)
NL, National League (1 of 2 that makes up major leagues)
PL, Player's League (short-lived in 19th centrury)
NA, National Association (1871-1875)
AA, American Association?
AL, American League (1 of 2 that makes up major leagues)
FL, Federal League (old major leagues, 1914)



3. create mlb.db and write four tables
```{r}
library(RSQLite)
conn <- dbConnect(SQLite(), 'mlb.db')

dbWriteTable(conn = conn, name = "game_log", value = game_log, row_names = FALSE, header = TRUE)
dbWriteTable(conn = conn, name = "park_codes", value = park_codes, row_names = FALSE, header = TRUE)
dbWriteTable(conn = conn, name = "person_codes", value = person_codes, row_names = FALSE, header = TRUE)
dbWriteTable(conn = conn, name = "team_codes", value = team_codes, row_names = FALSE, header = TRUE)

```

4. create game_id for game_log. column: h_name(3-digit), date (8 digit), number_of_game (1 digit)

```{r}
library(stringr)
game_id <- str_c(game_log$h_name, game_log$date, game_log$number_of_game)
head(game_id)
query1 <- 'ALTER TABLE game_log ADD COLUMN game_id TEXT;'
dbExecute(conn, query1)

query2 <- 'UPDATE game_log SET game_id = (h_name || CAST(date As Int) || CAST(number_of_game As Int));'
dbExecute(conn, query2)

dbGetQuery(conn, 'SELECT * FROM game_log LIMIT 5;')
dbGetQuery(conn, 'SELECT * FROM team_codes LIMIT 20;')
dbGetQuery(conn, 'SELECT * FROM park_codes LIMIT 5;')
dbGetQuery(conn, "SELECT DISTINCT(league) FROM park_codes;")
dbListTables(conn)
```
 5. create normalized tables
```{r}
query <- 'CREATE TABLE person (
person_id TEXT,
first_name TEXT,
last_name TEXT,
PRIMARY KEY (person_id)
);'
dbExecute(conn, query)

query <- 'INSERT INTO person
SELECT id person_id, 
first first_name, 
last last_name
FROM person_codes;'

dbExecute(conn, query)

#dbExecute(conn, "DROP TABLE person")
dbListTables(conn)

dbGetQuery(conn, "SELECT * FROM person LIMIT 5;")

query <- "CREATE TABLE park (
park_id TEXT,
name TEXT,
nickname TEXT,
city TEXT,
state TEXT,
notes TEXT,
PRIMARY KEY (park_id)
);"
dbExecute(conn, query)
query <-"INSERT INTO park
SELECT park_id,
name,
aka nickname,
city,
state,
notes
FROM park_codes;
"
dbExecute(conn, query)
dbGetQuery(conn, "SELECT * FROM park LIMIT 5;")


#league table
query <- "CREATE TABLE league (
league_id TEXT,
name TEXT,
PRIMARY KEY (league_id)
);"
dbExecute(conn, query)

query <- "INSERT INTO league VALUES 
('NL', 'National Leagues'),
('UA', 'Union Association'),
('AL', 'American League'),
('AA', 'American Association'),
('FL', 'Federal League'),
('PL', 'Player''s League');"

dbExecute(conn, query)
dbGetQuery(conn, "SELECT * FROM league;")

dbWriteTable(conn = conn, name = "appearance_type", value = "appearance_type.csv", row_names = FALSE, header = TRUE)

dbGetQuery(conn, "SELECT * FROM appearance_type;")
dbGetQuery(conn, "SELECT COUNT(*) FROM appearance_type;")


query <- "CREATE TABLE team (
team_id TEXT,
league_id TEXT,
city TEXT,
nickname TEXT,
franch_id TEXT,
PRIMARY KEY (team_id),
FOREIGN KEY (league_id) REFERENCES league(league_id)
);
"

#two team with the same team_id MIL
#Their time do not overlap, same nickname, so same team
query <- "UPDATE team_codes SET team_id = 'MIL2' WHERE (team_id = 'MIL') AND (league = 'NL');
"
dbExecute(conn, query)

query <- "INSERT INTO team
SELECT team_id,
league league_id,
city,
nickname,
franch_id
FROM team_codes;
"
dbExecute(conn, query)
dbGetQuery(conn, "SELECT * FROM team ORDER BY team_id;")

query <- "CREATE TABLE game (
game_id TEXT,
date DATETIME,
number_of_game INTEGER,
park_id TEXT,
length_outs INTEGER,
day INTEGER,
completion INTEGER,
forfeit TEXT,
protest INTEGER,
attendance INTEGER,
length_minutes INTEGER,
additional_info TEXT,
acquisition_info TEXT,
PRIMARY KEY (game_id),
FOREIGN KEY (park_id) REFERENCES park(park_id)
)
"
dbExecute(conn, query)
dbGetQuery(conn, "SELECT * FROM game_log LIMIT 1;")

query <- "INSERT INTO game
SELECT game_id,
date,
number_of_game,
park_id,
length_outs,
day_night day,
completion,
forefeit forfeit,
protest,
attendance,
length_minutes,
additional_info,
acquisition_info
FROM game_log;
"

query <- "UPDATE game SET day = 1 WHERE day = 'D';"
dbExecute(conn, query)
query <- "UPDATE game SET day = 0 WHERE day = 'N';"
query <- "SELECT COUNT(day) FROM game GROUP BY day"
dbGetQuery(conn, query)
dbGetQuery(conn, "SELECT * FROM game LIMIT 5;")


query <- 'SELECT sql FROM sqlite_master
WHERE name = "game_log"
  AND type = "table";
'
temp <- dbGetQuery(conn, query)

dbGetQuery(conn, "SELECT * FROM game_log LIMIT 5;")

library(stringr)
temp <- str_replace_all(temp,"[vh]_","")
temp <- str_replace_all(temp,"\\n ","")

query <- 
"CREATE TABLE `team_appearance` (
team_id TEXT,
game_id TEXT,
home INTEGER,
`league_id` TEXT, `score` REAL, 
`line_score` TEXT, `at_bats` REAL, `hits` REAL, `doubles` REAL, `triples` REAL, `homeruns` REAL, `rbi` REAL, `sacrifice_hits` REAL, `sacrifice_flies` REAL, `hit_by_pitch` REAL, 
`walks` REAL, `intentional_walks` REAL, `strikeouts` REAL, `stolen_bases` REAL, `caught_stealing` REAL, 
`grounded_into_double` REAL, `first_catcher_interference` INTEGER, `left_on_base` REAL, `pitchers_used` REAL, `individual_earned_runs` REAL, `team_earned_runs` REAL, `wild_pitches` REAL, `balks` REAL, `putouts` REAL, `assists` REAL, 
`errors` REAL, `passed_balls` REAL, `double_plays` REAL, `triple_plays` REAL,
PRIMARY KEY (team_id, game_id),
FOREIGN KEY (league_id) REFERENCES league(league_id));"
dbExecute(conn, query)


query <- "INSERT INTO team_appearance
SELECT h_name, game_id, 1 AS home, h_league, h_score, 
h_line_score, h_at_bats,h_hits, h_doubles, h_triples, 
h_homeruns, h_rbi, h_sacrifice_hits, h_sacrifice_flies, h_hit_by_pitch, 
h_walks, h_intentional_walks, h_strikeouts, h_stolen_bases, h_caught_stealing, h_grounded_into_double, h_first_catcher_interference, h_left_on_base, h_pitchers_used, h_individual_earned_runs, 
h_team_earned_runs, h_wild_pitches, h_balks, h_putouts, h_assists, 
h_errors, h_passed_balls, h_double_plays, h_triple_plays FROM game_log
UNION
SELECT v_name, game_id, 0 AS home, v_league, v_score, v_line_score, v_at_bats,
v_hits, v_doubles, v_triples, v_homeruns, v_rbi, v_sacrifice_hits, v_sacrifice_flies, v_hit_by_pitch, v_walks, v_intentional_walks, v_strikeouts, v_stolen_bases, v_caught_stealing, v_grounded_into_double, v_first_catcher_interference, v_left_on_base, v_pitchers_used, v_individual_earned_runs, 
v_team_earned_runs, v_wild_pitches, v_balks, v_putouts, v_assists, v_errors, v_passed_balls, v_double_plays, v_triple_plays FROM game_log;
"
dbExecute(conn, query)
dbGetQuery(conn, "SELECT * FROM team_appearance WHERE at_bats IS NOT NULL")


#person_appearance
query <- "CREATE TABLE person_appearance (
appearance_id TEXT,
person_id TEXT,
team_id TEXT,
game_id TEXT,
appearance_type_id TEXT,
PRIMARY KEY (appearance_id),
FOREIGN KEY (person_id) REFERENCES person(person_id),
FOREIGN KEY (team_id) REFERENCES team(team_id),
FOREIGN KEY (game_id) REFERENCES game(game_id),
FOREIGN KEY (appearance_type_id) REFERENCES appearance_type(appearance_type_id)
);
"
dbExecute(conn, query)

dbGetQuery(conn, "SELECT * FROM appearance_type;")
dbGetQuery(conn, "SELECT DISTINCT(v_player_1_def_pos) FROM game_log ;")
dbGetQuery(conn, "SELECT * FROM game_log WHERE saving_pitcher_id IS NOT NULL LIMIT 5 ;")

query <- "INSERT INTO person_appearance (
person_id,
team_id,
game_id,
appearance_type_id
)
SELECT hp_umpire_id, NULL AS team_id, game_id, 'MM' FROM game_log 
  WHERE hp_umpire_id IS NOT NULL
UNION
SELECT `1b_umpire_id`, NULL AS team_id, game_id, 'U1B' FROM game_log 
  WHERE `1b_umpire_id` IS NOT NULL
UNION
SELECT `2b_umpire_id`, NULL AS team_id, game_id, 'U2B' FROM game_log 
  WHERE `2b_umpire_id` IS NOT NULL
UNION
SELECT `3b_umpire_id`, NULL AS team_id, game_id, 'U3B' FROM game_log 
  WHERE `3b_umpire_id` IS NOT NULL
UNION
SELECT `lf_umpire_id`, NULL AS team_id, game_id, 'ULF' FROM game_log 
  WHERE `lf_umpire_id` IS NOT NULL
UNION
SELECT `rf_umpire_id`, NULL AS team_id, game_id, 'URF' FROM game_log 
  WHERE `rf_umpire_id` IS NOT NULL
UNION
SELECT `v_manager_id`, v_name, game_id, 'MM' FROM game_log 
  WHERE `v_manager_id` IS NOT NULL
UNION
SELECT `h_manager_id`, h_name, game_id, 'MM' FROM game_log 
  WHERE `h_manager_id` IS NOT NULL
UNION
SELECT winning_pitcher_id, 
        CASE 
          WHEN h_score > v_score THEN h_name
          ELSE v_name
          END,
       game_id, 'AWP' from game_log WHERE winning_pitcher_id IS NOT NULL
UNION
SELECT losing_pitcher_id,
      CASE
        WHEN h_score > v_score THEN v_name
        ELSE h_name
        END,
      game_id, 'ALP' from game_log WHERE losing_pitcher_id IS NOT NULL
 UNION
SELECT saving_pitcher_id,
       CASE
        WHEN h_score > v_score THEN h_name
        ELSE v_name
        END,
      game_id, 'ASP' from game_log WHERE saving_pitcher_id IS NOT NULL
UNION
SELECT winning_RBI_batter_id,
       CASE
        WHEN h_score > v_score THEN h_name
        ELSE v_name
        END,
      game_id, 'AWB' from game_log WHERE winning_RBI_batter_id IS NOT NULL ;
"
dbExecute(conn, query)

query <- "INSERT INTO person_appearance (
person_id,
team_id,
game_id,
appearance_type_id
)
SELECT v_starting_pitcher_id, v_name, game_id, 'PSP' from game_log WHERE v_starting_pitcher_id IS NOT NULL
UNION
SELECT h_starting_pitcher_id, h_name, game_id, 'PSP' from game_log WHERE h_starting_pitcher_id IS NOT NULL;
"
dbExecute(conn, query)

for (letter in c("h", "v")) {
  for (num in 1:9) {
    template <- '
      INSERT INTO person_appearance (
          game_id,
          team_id,
          person_id,
          appearance_type_id
      ) 
          SELECT
              game_id,
              %s_name,
              %s_player_%f_id,
              "O%f"
          FROM game_log
          WHERE %s_player_%f_id IS NOT NULL

      UNION

          SELECT
              game_id,
              %s_name,
              %s_player_%f_id,
              "D" || CAST(%s_player_%f_def_pos AS INT)
          FROM game_log
          WHERE %s_player_%f_id IS NOT NULL;
    '
    # replace all of the %s and %f with the correct letter number
    template <- gsub("%s", letter, template, fixed = TRUE)
    template <- gsub("%f", num, template, fixed = TRUE)

    dbExecute(conn, template)
  }
}

dbGetQuery(conn, "SELECT COUNT(*) FROM person_appearance")
dbGetQuery(conn, "SELECT * FROM person_appearance LIMIT 5;")
dbGetQuery(conn, "SELECT DISTINCT(h_player_1_def_pos) FROM game_log;")

dbExecute(conn, "DROP TABLE game_log; DROP TABLE park_codes; DROP TABLE person_codes; DROP TABLE team_codes;")
dbListTables(conn)

dbDisconnect(conn)
```
 
 